---
output: github_document
---


```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# dRiftDM

The package dRiftDM was developed to help psychological researchers apply and fit diffusion models to empirical data in the R environment. The most important feature is that dRiftDM can also handle non-stationary problems, that is, diffusion models with time-dependent parameters. The package comes with necessary tools for standard analyses, such as building a model, estimating parameters for multiple participants (separately per participant), or creating summary statistics. Pre-built models that are available within the package are: 

- The Standard (Ratcliff) Diffusion Model (Ratcliff, 1978, Psychological Review)
- The Diffusion Model for Conflict Tasks (Ulrich et al., 2015, Cognitive Psychology)
- The Shrinking Spotlight Model  (White et al., 2011, Cognitive Psychology)

Users can flexibly create other models and use the machinery of dRiftDM for estimating them. 

With Version 0.2.0 model predictions (i.e., their first-passage time) are derived by numerically solving the Kolmogorov-Forward-Equation or a coupled set of integral equations, based on code provided by Richter et al. (2023, Journal of Mathematical Psychology. 


## Notes

The current version is 0.2.0. In contrast to the former version 0.1.1, version 0.2.0 leverages more strongly the S3 object system. We also introduced a new way to handle parameters across conditions using so-called "flex_prms" objects. 

If you want install the old version 0.1.1. You can use: 

```{r, eval = F}
devtools::install_github("bucky2177/dRiftDM", ref = "0.1.1")
```

## Installation

You can install the (current) development version of dRiftDM from [GitHub](https://github.com/) with:

```{r, eval = F}
# install.packages("devtools")
devtools::install_github("bucky2177/dRiftDM")
```

## Example

We are in the progress of publishing a tutorial with more detailed instructions (see the respective [OSF pre-print](https://osf.io/preprints/osf/3t2vf). For now, here is an example on how to fit the Diffusion Model for Conflict Tasks to some synthetic data using bounded Nelder-Mead:

```{r example}
library(dRiftDM)
dmc_model = dmc_dm(dx = .002, dt = .002, t_max = 1.2)
obs_data(dmc_model) = dmc_synth_data
print(dmc_model)
```

```{r, eval = F}
est_model = estimate_model(
  drift_dm_obj = dmc_model,                       # a model
  lower = c(2, 0.4, 0.2, 0.005, 0.02, 0.02, 2),   # lower boundary - search space
  upper = c(6, 0.8, 0.5, 0.050, 0.12, 0.30, 7),   # upper boundary - search space
  verbose = 2,                                    # print out each evaluation
  use_de_optim = F, use_nmkb = T # use Nelder-Mead for demonstration
)
```


